#!/usr/bin/env Rscript

```{r}
# Load required libraries
library(dplyr)
library(readr)
library(here)
```
```{r}
# Read the FAOSTAT data
data <- read_csv(here("data", "FAOSTAT_data_en_4-28-2026.csv"))
```

# Display original data structure
cat("Original data structure:\n")
print(str(data))
cat("\n")
cat("Number of rows in original data:", nrow(data), "\n")
cat("Number of unique countries:", length(unique(data$Area)), "\n")
cat("Number of unique items:", length(unique(data$Item)), "\n\n")

# Aggregate data by Item, summing production values globally
global_production <- data %>%
  group_by(Item, `Item Code`, Year, Unit, Element, `Element Code`) %>%
  summarise(
    Global_Production = sum(Value, na.rm = TRUE),
    Number_of_Countries = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(Global_Production))

# Display summary of aggregated data
cat("Aggregated global production data:\n")
print(head(global_production, 20))
cat("\n")
cat("Number of rows in aggregated data:", nrow(global_production), "\n")

# Save the aggregated data to a new CSV file
output_file <- "/home/claude/global_production_by_item.csv"
write_csv(global_production, output_file)
cat("\nGlobal production data saved to:", output_file, "\n")

# Create a summary showing top 10 items by production
cat("\nTop 10 items by global production:\n")
top_items <- global_production %>%
  slice_max(order_by = Global_Production, n = 10) %>%
  select(Item, Year, Global_Production, Unit, Number_of_Countries)
print(top_items)