#!/usr/bin/env Rscript


#Part 1 - load and aggreagate data (note added new column mannually to output file labelling items as synthetic or not)
```{r}
# Load required libraries
library(dplyr)
library(readr)
#set here. Make sure to set the working directory to the root of the project first
library(here)
library(ggplot2)
library(scales)
library(readr)
library(here)
library(forcats)
library(stringr)
library(RColorBrewer)
library(tidyverse)
```
```{r}
# Read the FAOSTAT data
data <- read_csv(here("raw", "FAOSTAT_data_en_4-28-2026.csv"))
```

```{r}
# Display original data structure
cat("Original data structure:\n")
print(str(data))
cat("\n")
cat("Number of rows in original data:", nrow(data), "\n")
cat("Number of unique countries:", length(unique(data$Area)), "\n")
cat("Number of unique items:", length(unique(data$Item)), "\n\n")
```

```{r}
# Aggregate data by Item, summing production values globally
global_production <- data %>%
  group_by(Item, `Item Code`, Year, Unit, Element, `Element Code`) %>%
  summarise(
    Global_Production = sum(Value, na.rm = TRUE),
    Number_of_Countries = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(Global_Production))

# Display summary of aggregated data
cat("Aggregated global production data:\n")
print(head(global_production, 20))
cat("\n")
cat("Number of rows in aggregated data:", nrow(global_production), "\n")
```

```{r}
# Save the aggregated data to a new CSV file
output_file <- here("raw", "cleaned", "global_production_by_item.csv")
write_csv(global_production, output_file)
cat("\nGlobal production data saved to:", output_file, "\n")
```

```{r}
# Create a summary showing top 10 items by production
cat("\nTop 10 items by global production:\n")
top_items <- global_production %>%
  slice_max(order_by = Global_Production, n = 10) %>%
  select(Item, Year, Global_Production, Unit, Number_of_Countries)
print(top_items)
```

# Part 2 - create a plot of the items by production

## colour schemes
```{r}
# Define custom colors
my_col_1 <- c("#440154FF", "#470E61FF", "#481B6DFF", "#482576FF", "#46307EFF", "#443B84FF", "#404688FF", "#3C508BFF", "#38598CFF", "#33628DFF", "#2F6B8EFF", "#2C738EFF", "#287C8EFF", "#25838EFF", "#228C8DFF", "#1F948CFF", "#1E9D89FF", "#20A486FF", "#26AD81FF", "#31B57BFF", "#3FBC73FF", "#4FC46AFF", "#61CB5FFF", "#75D054FF", "#8BD646FF", "#A2DA37FF", "#B9DE28FF", "#D1E11CFF", "#E8E419FF", "#FDE725FF")

my_col_2 <- c("#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D", "#666666", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928", "#E41A1C", "#999999")

my_col_3 <- c("#999999", "#E41A1C", "#B15928", "#FFFF99", "#6A3D9A", "#CAB2D6", "#FF7F00", "#FDBF6F", "#E31A1C", "#FB9A99", "#33A02C", "#B2DF8A", "#1F78B4", "#A6CEE3", "#666666", "#A6761D", "#E6AB02", "#66A61E", "#E7298A", "#7570B3", "#D95F02", "#1B9E77", "#B3B3B3", "#E5C494", "#FFD92F", "#A6D854", "#E78AC3", "#8DA0CB", "#FC8D62", "#66C2A5")
```

## a pie chart of the items by global production
```{r}


# Read the manually modified data
cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# Order the data by Global_Production (highest to lowest)
# and convert Item to an ordered factor
cleaned_data <- cleaned_data %>%
  arrange(desc(Global_Production)) %>%
  mutate(Item = factor(Item, levels = Item))

# Create a pie chart of the items by global production
pie_chart <- ggplot(cleaned_data, aes(x = "", y = Global_Production, fill = Item)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Items by Global Production") +
  scale_fill_manual(values = my_col_2)

print(pie_chart)
```

## a pie chart of the items by global production with small items (< 2%) grouped into "Other"
```{r}


# Read the data
cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# Calculate percentage of total
cleaned_data <- cleaned_data %>%
  mutate(Percentage = Global_Production / sum(Global_Production) * 100)

# Group items < 2% into "Other"
threshold <- 2  # You can adjust this threshold

cleaned_data_grouped <- cleaned_data %>%
  mutate(Item_Display = ifelse(Percentage < threshold, "Other", Item)) %>%
  group_by(Item_Display) %>%
  summarise(Global_Production = sum(Global_Production)) %>%
  arrange(desc(Global_Production)) %>%
  mutate(Item_Display = factor(Item_Display, levels = Item_Display))

# Create pie chart
pie_chart <- ggplot(cleaned_data_grouped, aes(x = "", y = Global_Production, fill = Item_Display)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Items by Global Production (Small items grouped)") +
  scale_fill_manual(values = my_col_2)

print(pie_chart)
```


## a horizontal bar chart of the items by global production

```{r}
cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# 1. Abbreviate labels and reorder
cleaned_data <- cleaned_data %>%
  mutate(Item_clean = case_when(
    Item == "Urea" ~ "Urea",
    Item == "Potassium chloride (muriate of potash) (MOP)" ~ "MOP",
    Item == "NPK fertilizers" ~ "NPK fertilizers",
    Item == "Ammonium nitrate (AN)" ~ "AN",
    Item == "Other NP compounds" ~ "Other NP",
    Item == "Ammonia, anhydrous" ~ "Anhydrous ammonia",
    Item == "Diammonium phosphate (DAP)" ~ "DAP",
    Item == "Superphosphates, other" ~ "Other superphosphates",
    Item == "Ammonium sulphate" ~ "Ammonium sulphate",
    Item == "Potassium nitrate" ~ "Potassium nitrate",
    Item == "Phosphate rock" ~ "Phosphate rock",
    Item == "Other phosphatic fertilizers, n.e.c." ~ "Other P fert.",
    Item == "Calcium ammonium nitrate (CAN) and other mixtures with calcium carbonate" ~ "CAN + CaCO3",
    Item == "Urea and ammonium nitrate solutions (UAN)" ~ "UAN",
    Item == "Monoammonium phosphate (MAP)" ~ "MAP",
    Item == "Fertilizers n.e.c." ~ "Fertilizers n.e.c.",
    Item == "Superphosphates above 35%" ~ "Superphos. above 35%",
    Item == "Other nitrogenous fertilizers, n.e.c." ~ "Other N fert.",
    Item == "Potassium sulphate (sulphate of potash) (SOP)" ~ "SOP",
    Item == "PK compounds" ~ "PK compounds",
    Item == "Other potassic fertilizers, n.e.c." ~ "Other K fert.",
    Item == "Sodium nitrate" ~ "Sodium nitrate",
    Item == "Other NK compounds" ~ "Other NK",
    TRUE ~ str_trunc(Item, 18)
  )) %>%
  mutate(Item_clean = fct_reorder(Item_clean, Global_Production)) %>%
  mutate(label_pos = ifelse(Global_Production > max(Global_Production) * 0.3,
                            "inside", "outside"))

# 2. Use your original palette - assuming my_col_2 exists

bar_chart <- ggplot(cleaned_data, aes(x = Global_Production, y = Item_clean)) +
  geom_col(fill = "#A2DA37FF", color = "white", linewidth = 0.2) +
  # Labels inside bars - now black
  geom_text(data = filter(cleaned_data, label_pos == "inside"),
            aes(label = label_number(scale = 1e-6, accuracy = 0.0001)(Global_Production)),
            hjust = 1.1, size = 3, color = "black", fontface = "bold") +
  # Labels outside bars  
  geom_text(data = filter(cleaned_data, label_pos == "outside"),
            aes(label = label_number(scale = 1e-6, accuracy = 0.0001)(Global_Production)),
            hjust = -0.1, size = 3, color = "grey20") +
  theme_minimal(base_size = 11) +
  labs(x = "Global Production (Mt)", y = NULL, title = NULL) + 
  scale_x_continuous(
    labels = label_number(scale = 1e-6),
    expand = expansion(mult = c(0, 0.18))
  ) +
  theme(
    legend.position = "none", # hide legend since each bar = one item
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9, color = "grey20"),
    axis.text.x = element_text(size = 9, color = "grey20"),
    axis.title.x = element_text(size = 10, color = "grey20", margin = margin(t = 8)),
    plot.margin = margin(0, 0, 0, 0)
  )

print(bar_chart)
```




## a pie chart of the items by production category (synthetic, non-synthetic, blended, unspecified) - note this requires a new column to be added to the data manually labelling each item as synthetic or not
```{r}
# 1. Load data
cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# --- Prep data ---
plot_data <- cleaned_data %>%
  group_by(Category) %>%
  summarise(total_production = sum(Global_Production, na.rm = TRUE)) %>%
  mutate(
    fraction = total_production / sum(total_production),
    pct = fraction * 100,
    pct_label = paste0(round(pct, 1), "%")
  ) %>%
  arrange(desc(fraction)) %>% 
  mutate(
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n=-1)),
    label_y = (ymax + ymin) / 2,
    label_x = ifelse(Category %in% c("Blended", "Unspecified"), 2.3, 1.5)
  )

# Lock factor order
plot_data$Category <- factor(plot_data$Category, levels = plot_data$Category)

category_colors <- c(
  "Synthetic" = "#66C2A5",      
  "Unspecified" = "#FC8D62",    
  "Non-synthetic" = "#8DA0CB",  
  "Blended" = "#E78AC3"        
)

# --- Donut chart ---
pie_chart <- ggplot(plot_data, aes(ymin = ymin, ymax = ymax, xmin = 1, xmax = 2, fill = Category)) +
  geom_rect(color = "white", linewidth = 0.3) +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(x = label_x, y = label_y, label = pct_label),
            size = 2.8, fontface = "bold") +
  geom_segment(data = plot_data %>% filter(Category %in% c("Blended", "Unspecified")),
               aes(x = 2, xend = 2.2, y = label_y, yend = label_y),
               color = "grey30", linewidth = 0.3) +
  scale_fill_manual(values = category_colors) +
  xlim(0.6, 2.4) + 
  labs(
    title = paste("Total:", format(round(sum(plot_data$total_production), 0), big.mark = ","), "tonnes"),
    fill = NULL
  ) +
  theme_void(base_size = 8) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 7, margin = margin(b = 2)),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.key.size = unit(0.25, "in"),
    legend.key.width = unit(0.25, "in"),
    legend.text = element_text(size = 6),
    legend.spacing.x = unit(0.1, "in"), # fixes legend smooshing
    legend.margin = margin(t = -5),
    plot.margin = margin(0, 0, 0, 0)
  ) +
  guides(fill = guide_legend(ncol = 1, bycol = TRUE))

print(pie_chart)
```
    
```{r}
bar_svg_path <- here("analysis", "global_production_bar_chart.svg")
pie_svg_path <- here("analysis", "synthetic_pie_chart.svg")
pdf_dir  <- path.expand("~/writing/bbro/lit_review/images")
# 1. Save SVGs
ggsave(bar_svg_path, plot = bar_chart, width = 4, height = 5, units = "in", dpi = 600)
ggsave(pie_svg_path, plot = pie_chart, width = 3, height = 6, units = "in", dpi = 600)

# 2. Change to target dir so pdf_tex lands there
old_wd <- getwd()
setwd(pdf_dir)
on.exit(setwd(old_wd))

# 3. Function to export one SVG via Inkscape
export_svg_to_pdftex <- function(svg_file, pdf_name) {
  system2(
    command = "inkscape",
    args = c(
      svg_file,
      "--export-area-drawing",
      "--export-type=pdf",
      "--export-latex",
      paste0("--export-filename=", pdf_name)
    )
  )
}

# 4. Export both
export_svg_to_pdftex(bar_svg_path, "global_production_bar_chart.pdf")
export_svg_to_pdftex(pie_svg_path, "synthetic_pie_chart.pdf")

cat("\nFiles created:\n")

```