#!/usr/bin/env Rscript


#Part 1 - load and aggreagate data (note added new column mannually to output file labelling items as synthetic or not)
```{r}
# Load required libraries
library(dplyr)
library(readr)
```
```{r}
#set here
here::iam("analysis/global_fert_production.rmd")

# Read the FAOSTAT data
data <- read_csv(here("raw", "FAOSTAT_data_en_4-28-2026.csv"))
```

```{r}
# Display original data structure
cat("Original data structure:\n")
print(str(data))
cat("\n")
cat("Number of rows in original data:", nrow(data), "\n")
cat("Number of unique countries:", length(unique(data$Area)), "\n")
cat("Number of unique items:", length(unique(data$Item)), "\n\n")
```

```{r}
# Aggregate data by Item, summing production values globally
global_production <- data %>%
  group_by(Item, `Item Code`, Year, Unit, Element, `Element Code`) %>%
  summarise(
    Global_Production = sum(Value, na.rm = TRUE),
    Number_of_Countries = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(Global_Production))

# Display summary of aggregated data
cat("Aggregated global production data:\n")
print(head(global_production, 20))
cat("\n")
cat("Number of rows in aggregated data:", nrow(global_production), "\n")
```

```{r}
# Save the aggregated data to a new CSV file
output_file <- here("raw", "cleaned", "global_production_by_item.csv")
write_csv(global_production, output_file)
cat("\nGlobal production data saved to:", output_file, "\n")
```

```{r}
# Create a summary showing top 10 items by production
cat("\nTop 10 items by global production:\n")
top_items <- global_production %>%
  slice_max(order_by = Global_Production, n = 10) %>%
  select(Item, Year, Global_Production, Unit, Number_of_Countries)
print(top_items)
```

# Part 2 - create a plot of the items by production

## a pie chart of the items by global production
```{r}
library(ggplot2)

# Read the manually modified data
cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# Order the data by Global_Production (highest to lowest)
# and convert Item to an ordered factor
cleaned_data <- cleaned_data %>%
  arrange(desc(Global_Production)) %>%
  mutate(Item = factor(Item, levels = Item))

# Define custom colors
my_col_1 <- c("#440154FF", "#470E61FF", "#481B6DFF", "#482576FF", "#46307EFF", "#443B84FF", "#404688FF", "#3C508BFF", "#38598CFF", "#33628DFF", "#2F6B8EFF", "#2C738EFF", "#287C8EFF", "#25838EFF", "#228C8DFF", "#1F948CFF", "#1E9D89FF", "#20A486FF", "#26AD81FF", "#31B57BFF", "#3FBC73FF", "#4FC46AFF", "#61CB5FFF", "#75D054FF", "#8BD646FF", "#A2DA37FF", "#B9DE28FF", "#D1E11CFF", "#E8E419FF", "#FDE725FF")

my_col_2 <- c("#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D", "#666666", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928", "#E41A1C", "#999999")
# Create a pie chart of the items by global production
pie_chart <- ggplot(cleaned_data, aes(x = "", y = Global_Production, fill = Item)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Items by Global Production") +
  scale_fill_manual(values = my_col_2)

print(pie_chart)
```

## a pie chart of the items by global production with small items (< 2%) grouped into "Other"
```{r}
library(ggplot2)
library(dplyr)

# Read the data
cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# Calculate percentage of total
cleaned_data <- cleaned_data %>%
  mutate(Percentage = Global_Production / sum(Global_Production) * 100)

# Group items < 2% into "Other"
threshold <- 2  # You can adjust this threshold

cleaned_data_grouped <- cleaned_data %>%
  mutate(Item_Display = ifelse(Percentage < threshold, "Other", Item)) %>%
  group_by(Item_Display) %>%
  summarise(Global_Production = sum(Global_Production)) %>%
  arrange(desc(Global_Production)) %>%
  mutate(Item_Display = factor(Item_Display, levels = Item_Display))

# Create pie chart
pie_chart <- ggplot(cleaned_data_grouped, aes(x = "", y = Global_Production, fill = Item_Display)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Items by Global Production (Small items grouped)") +
  scale_fill_manual(values = my_col_2)

print(pie_chart)
```


## a horizontal bar chart of the items by global production
```{r}
library(ggplot2)
library(dplyr)
library(scales)

cleaned_data <- read_csv(here("raw", "cleaned", "global_production_by_item_modified.csv"))

# Order by production
cleaned_data <- cleaned_data %>%
  arrange(desc(Global_Production)) %>%
  mutate(Item = factor(Item, levels = Item)) %>%
  mutate(label_position = ifelse(Global_Production > max(Global_Production) * 0.1, 
                                  "inside", "outside"))

# Create horizontal bar chart
bar_chart <- ggplot(cleaned_data, aes(x = Global_Production, y = Item, fill = Item)) +
  geom_col(color = "white") +
  geom_text(data = filter(cleaned_data, label_position == "inside"),
            aes(label = comma(Global_Production)), 
            hjust = 1.1, size = 3, color = "white", fontface = "bold") +
  geom_text(data = filter(cleaned_data, label_position == "outside"),
            aes(label = comma(Global_Production)), 
            hjust = -0.1, size = 3, color = "black") +
  theme_minimal() +
  labs(title = "Global Production by Fertilizer Type",
       x = "Global Production (tonnes)",
       y = NULL) +
  scale_fill_manual(values = my_col_2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 9))

print(bar_chart)
```