---
title: "DEG Visualisation - Sleuth and Rank Products"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Load libraries
```{r}
library(tidyverse)
library(here)
library(ggplot2)
library(patchwork)
library(scales)
```

# Output directory
```{r}
fig_dir <- "C:/Users/ges23jir/Documents/Coding/TSL/mst7_transcriptome/Analysis/Data/Figures/DEG_analysis"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

save_plot <- function(plot, filename, width = 10, height = 7) {
  ggsave(file.path(fig_dir, paste0(filename, ".pdf")),
         plot, width = width, height = height, device = "pdf")
  ggsave(file.path(fig_dir, paste0(filename, ".svg")),
         plot, width = width, height = height, device = "svg")
  message("Saved: ", filename)
}
```

# ── SLEUTH RESULTS ────────────────────────────────────────────────────────────

## Load all sleuth CSVs
```{r}
dge_dir <- here("Analysis/Data/DGE")

sleuth_files <- list.files(dge_dir, pattern = "^sleuth_.*\\.csv$", full.names = TRUE)

sleuth_all <- lapply(sleuth_files, function(f) {
  df <- read_csv(f, col_types = cols())
  
  # Parse filename: sleuth_Exp1_8H_MST7WT_vs_Guy11.csv
  bn <- basename(f) %>% gsub("^sleuth_|.csv$", "", .)
  parts <- str_split(bn, "_")[[1]]
  
  df %>% mutate(
    experiment  = parts[1],
    timepoint   = parts[2],
    condition   = parts[3],
    control     = parts[5],
    comparison  = paste0(parts[3], "_vs_", parts[5]),
    source      = "sleuth"
  )
}) %>% bind_rows()

# Classify direction using b (beta = log2 fold change) and qval
sleuth_all <- sleuth_all %>%
  mutate(
    sig       = !is.na(qval) & qval < 0.05,
    direction = case_when(
      sig & b > 0 ~ "up",
      sig & b < 0 ~ "down",
      TRUE        ~ "ns"
    ),
    timepoint_num = as.numeric(gsub("H", "", timepoint))
  )
```

## Bar chart: DEG counts per comparison per timepoint (sleuth)
```{r}
sleuth_counts <- sleuth_all %>%
  filter(direction != "ns") %>%
  count(experiment, comparison, timepoint, timepoint_num, direction) %>%
  mutate(
    # Upregulated goes right (positive), downregulated goes left (negative)
    n_plot    = if_else(direction == "down", -n, n),
    # Reverse order so 0H is at top, 24H at bottom
    timepoint = fct_reorder(timepoint, -timepoint_num)
  )

comparisons <- unique(sleuth_counts$comparison)

plots_sleuth <- lapply(comparisons, function(comp) {
  df <- sleuth_counts %>% filter(comparison == comp)
  exp_label <- unique(df$experiment)
  
  ggplot(df, aes(y = timepoint, x = n_plot, fill = direction)) +
    geom_col(width = 0.6) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    scale_fill_manual(values = c(up = "#D62728", down = "#1F77B4"),
                      labels = c(up = "Upregulated", down = "Downregulated")) +
    scale_x_continuous(labels = function(x) abs(x)) +
    labs(
      title    = paste0("DEGs: ", gsub("_vs_", " vs ", comp)),
      subtitle = paste0(exp_label, " | sleuth | qval < 0.05"),
      y        = "Timepoint",
      x        = "Number of DEGs",
      fill     = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title      = element_text(face = "bold"),
      legend.position = "bottom"
    )
})
names(plots_sleuth) <- comparisons

# Save individual comparison plots
for (comp in comparisons) {
  save_plot(plots_sleuth[[comp]],
            paste0("sleuth_DEG_barplot_", comp))
}
```

## Combined overview: all sleuth comparisons faceted
```{r}
p_sleuth_overview <- ggplot(sleuth_counts,
                            aes(y = timepoint, x = n_plot, fill = direction)) +
  geom_col(width = 0.6) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  scale_fill_manual(values = c(up = "#D62728", down = "#1F77B4"),
                    labels = c(up = "Up", down = "Down")) +
  scale_x_continuous(labels = function(x) abs(x)) +
  facet_wrap(~ comparison, scales = "free_x", ncol = 3) +
  labs(
    title = "DEG counts across all sleuth comparisons",
    y     = "Timepoint",
    x     = "Number of DEGs (qval < 0.05)",
    fill  = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.text      = element_text(face = "bold", size = 8),
    legend.position = "bottom"
  )

save_plot(p_sleuth_overview, "sleuth_DEG_overview_all_comparisons",
          width = 16, height = 12)
```

## Volcano plots for each sleuth comparison × timepoint
```{r}
volcano_sleuth <- function(df, comp, tp) {
  df <- df %>%
    filter(comparison == comp, timepoint == tp) %>%
    mutate(
      log10q   = -log10(pmax(qval, 1e-30)),
      label    = if_else(sig & abs(b) > 1, target_id, NA_character_)
    )
  
  if (nrow(df) == 0) return(NULL)
  
  ggplot(df, aes(x = b, y = log10q, colour = direction)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linewidth = 0.3) +
    scale_colour_manual(values = c(up = "#D62728", down = "#1F77B4", ns = "grey70"),
                        guide  = "none") +
    labs(
      title    = paste0(gsub("_vs_", " vs ", comp), " | ", tp),
      subtitle = "sleuth | qval < 0.05",
      x        = "Beta (log2 fold change)",
      y        = "-log10(qval)"
    ) +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))
}

for (comp in comparisons) {
  tps <- unique(sleuth_all$timepoint[sleuth_all$comparison == comp])
  vol_plots <- lapply(tps, function(tp) volcano_sleuth(sleuth_all, comp, tp))
  vol_plots <- Filter(Negate(is.null), vol_plots)
  
  if (length(vol_plots) == 0) next
  
  combined <- wrap_plots(vol_plots, ncol = min(3, length(vol_plots)))
  save_plot(combined,
            paste0("sleuth_volcano_", comp),
            width  = 5 * min(3, length(vol_plots)),
            height = 5 * ceiling(length(vol_plots) / 3))
}
```

# ── RANK PRODUCTS RESULTS ─────────────────────────────────────────────────────

## Load all rank products CSVs
```{r}
rp_dir <- here("Analysis/Data/DGE/RankProducts")

rp_files <- list.files(rp_dir, pattern = "^rankprod_.*\\.csv$", full.names = TRUE)

rp_all <- lapply(rp_files, function(f) {
  df <- read_csv(f, col_types = cols())
  bn <- basename(f) %>% gsub("^rankprod_|\\.csv$", "", .)
  
  # Extract components strictly
  tp_str <- sub("_.*$", "", bn)
  # Everything after the timepoint: e.g., "Guy11_vs_Guy11M"
  comp_str <- sub("^[^_]+_", "", bn) 
  
  # Extract grp1 and grp2 from the filename itself
  parts <- str_split(comp_str, "_vs_")[[1]]
  g1 <- parts[1]
  g2 <- parts[2]
  
  df %>% mutate(
    timepoint = tp_str,
    comparison = comp_str,
    grp1_name = g1,
    grp2_name = g2,
    source = "rankprod"
  )
}) %>% bind_rows()

rp_all <- rp_all %>%
  mutate(
    sig = !is.na(pfp) & pfp < 0.05,
    timepoint_num = as.numeric(gsub("H", "", timepoint))
  )
```

## Bar chart: DEG counts per rank products comparison per timepoint
```{r}
save_plot <- function(plot, filename, width = 10, height = 7) {
  # Ensure fig_dir exists
  if(!exists("fig_dir")) fig_dir <- "Analysis/Data/Figures/DEG_analysis"
  
  # Save as PDF
  ggsave(file.path(fig_dir, paste0(filename, ".pdf")),
         plot = plot, width = width, height = height, device = "pdf")
  
  # Save as SVG
  ggsave(file.path(fig_dir, paste0(filename, ".svg")),
         plot = plot, width = width, height = height, device = "svg")
  
  message("Saved: ", filename)
}

# Assuming rp_all is already loaded as per your previous steps
rp_comparisons <- unique(rp_all$comparison)

plots_rp <- lapply(rp_comparisons, function(comp) {
  
  # 1. Identify Group 1 and Group 2 from the filename/comparison string
  # Example: "pmk1_vs_mst7" -> g1 is "pmk1", g2 is "mst7"
  parts <- strsplit(comp, "_vs_")[[1]]
  g1 <- parts[1]
  g2 <- parts[2]
  
  df_plot <- rp_all %>%
    filter(comparison == comp, sig) %>%
    mutate(
      # 2. Assign direction BASED ON THE FILENAME ORDER
      # If the label is "up_in_[First Group]", it's "up" (Positive x-axis)
      # If the label is "up_in_[Second Group]", it's "down" (Negative x-axis)
      dir_simple = case_when(
        direction == paste0("up_in_", g1) ~ "up",
        direction == paste0("up_in_", g2) ~ "down",
        TRUE ~ NA_character_
      ),
      timepoint_num = as.numeric(gsub("H", "", timepoint))
    ) %>%
    filter(!is.na(dir_simple)) %>% # Remove any mismatched labels
    count(timepoint, timepoint_num, dir_simple) %>%
    mutate(
      # 3. Mirror the bars: "down" goes to the left
      n_plot = if_else(dir_simple == "down", -n, n),
      timepoint = fct_reorder(timepoint, -timepoint_num)
    )
  
  if (nrow(df_plot) == 0) return(NULL)
  
  # 4. Create the plot
  ggplot(df_plot, aes(y = timepoint, x = n_plot, fill = dir_simple)) +
    geom_col(width = 0.6, color = "white", linewidth = 0.1) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    scale_fill_manual(
      values = c(up = "#D62728", down = "#1F77B4"),
      labels = c(up = paste0("Higher in ", g1), 
                 down = paste0("Higher in ", g2))
    ) +
    scale_x_continuous(labels = function(x) abs(x)) +
    labs(
      title    = paste0("DEGs: ", gsub("_vs_", " vs ", comp)),
      subtitle = "Rank Products | pfp < 0.05",
      x        = "Number of DEGs",
      y        = "Timepoint",
      fill     = "Expression Direction"
    ) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
})

names(plots_rp) <- rp_comparisons

# Save them again
for (comp in names(plots_rp)) {
  if (!is.null(plots_rp[[comp]])) {
    save_plot(plots_rp[[comp]], paste0("rankprod_DEG_barplot_", comp))
  }
}
```

## Combined overview: rank products
```{r}
rp_counts_all <- rp_all %>%
  filter(sig) %>%
  mutate(
    timepoint_num = as.numeric(gsub("H", "", timepoint)),
    # Extract the target group for every row based on its comparison name
    target_grp = sub("_vs_.*", "", comparison),
    dir_simple = if_else(direction == paste0("up_in_", target_grp), "up", "down")
  ) %>%
  count(comparison, timepoint, timepoint_num, dir_simple) %>%
  mutate(
    n_plot    = if_else(dir_simple == "down", -n, n),
    timepoint = fct_reorder(timepoint, -timepoint_num)
  )

if (nrow(rp_counts_all) > 0) {
  p_rp_overview <- ggplot(rp_counts_all,
                          aes(y = timepoint, x = n_plot, fill = dir_simple)) +
    geom_col(width = 0.6) +
    geom_vline(xintercept = 0, linewidth = 0.3) +
    scale_fill_manual(values = c(up = "#D62728", down = "#1F77B4"),
                      labels = c(up = "Up (First Group)", down = "Down (First Group)")) +
    scale_x_continuous(labels = function(x) abs(x)) +
    facet_wrap(~ comparison, scales = "free_x") +
    labs(
      title = "DEG counts across all Rank Products comparisons",
      y     = "Timepoint",
      x     = "Number of DEGs (pfp < 0.05)",
      fill  = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(strip.text = element_text(face = "bold"),
          legend.position = "bottom")
  
 # If you are running the combined overview code:
if (exists("p_rp_overview")) {
  ggsave(
    filename = file.path(fig_dir, "rankprod_DEG_overview.png"),
    plot = p_rp_overview,
    width = 14, 
    height = 8,
    dpi = 300
  )
}
}
```

## Summary tables: DEG counts per comparison × timepoint
```{r}
library(knitr)

# ── SLEUTH SUMMARY TABLE ──────────────────────────────────────────────────────
sleuth_summary <- sleuth_all %>%
  filter(sig) %>%
  mutate(timepoint = fct_reorder(timepoint, timepoint_num)) %>%
  count(comparison, timepoint, direction) %>%
  pivot_wider(
    names_from  = direction,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(total = up + down) %>%
  arrange(comparison, timepoint)

kable(sleuth_summary,
      caption = "Sleuth DEG counts per comparison and timepoint (qval < 0.05)",
      col.names = c("Comparison", "Timepoint", "Up", "Down", "Total"))

write_csv(sleuth_summary,
          file.path(fig_dir, "DEG_summary_table_sleuth.csv"))

# ── RANK PRODUCTS SUMMARY TABLE ───────────────────────────────────────────────
rp_summary <- rp_all %>%
  filter(sig) %>%
  mutate(
    timepoint_num = as.numeric(gsub("H", "", timepoint)),
    timepoint     = fct_reorder(timepoint, timepoint_num),
    # Use sub to get the first condition in the comparison string
    target_grp    = sub("_vs_.*", "", comparison),
    dir_simple    = if_else(direction == paste0("up_in_", target_grp), "up", "down")
  ) %>%
  count(comparison, timepoint, dir_simple) %>%
  pivot_wider(
    names_from  = dir_simple,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(total = up + down) %>%
  arrange(comparison, timepoint)

kable(rp_summary,
      caption = "Rank Products DEG counts (Up/Down relative to first group in comparison name)",
      col.names = c("Comparison", "Timepoint", "Up", "Down", "Total"))
```

# ── COMBINED SUMMARY: sleuth + rank products side by side ─────────────────────
```{r}
# For the direct comparisons that exist in both (pmk1_vs_mst7, Guy11_vs_Guy11M)
# compare DEG counts from both methods at each timepoint

shared_comparisons <- intersect(
  unique(sleuth_all$comparison),
  unique(rp_all$comparison)
)

if (length(shared_comparisons) > 0) {
  
  sleuth_shared <- sleuth_all %>%
    filter(comparison %in% shared_comparisons, direction != "ns") %>%
    count(comparison, timepoint, timepoint_num, direction) %>%
    mutate(method = "sleuth",
           n_plot = if_else(direction == "down", -n, n),
           timepoint = fct_reorder(timepoint, -timepoint_num))
  
  rp_shared <- rp_all %>%
    filter(comparison %in% shared_comparisons, sig) %>%
    mutate(direction = if_else(str_detect(direction, condition), "up", "down")) %>%
    count(comparison, timepoint, timepoint_num, direction) %>%
    mutate(method = "RankProd",
           n_plot = if_else(direction == "down", -n, n),
           timepoint = fct_reorder(timepoint, -timepoint_num))
  
  combined_shared <- bind_rows(sleuth_shared, rp_shared)
  
  p_method_compare <- ggplot(combined_shared,
                              aes(y = timepoint, x = n_plot,
                                  fill = direction, alpha = method)) +
    geom_col(position = "dodge", width = 0.7) +
    geom_vline(xintercept = 0, linewidth = 0.3) +
    scale_fill_manual(values  = c(up = "#D62728", down = "#1F77B4"),
                      labels  = c(up = "Up", down = "Down")) +
    scale_alpha_manual(values = c(sleuth = 1, RankProd = 0.6)) +
    scale_x_continuous(labels = function(x) abs(x)) +
    facet_wrap(~ comparison, scales = "free_x") +
    labs(
      title    = "DEG counts: sleuth vs Rank Products",
      subtitle = "Shared direct comparisons",
      y        = "Timepoint",
      x        = "Number of DEGs",
      fill     = NULL,
      alpha    = "Method"
    ) +
    theme_classic(base_size = 11) +
    theme(strip.text = element_text(face = "bold"),
          legend.position = "bottom")
  
  save_plot(p_method_compare, "method_comparison_sleuth_vs_rankprod",
            width = 12, height = 6)
}

message("All DEG visualisation plots saved to: ", fig_dir)
```
