---
title: "GO Term Analysis - MF and BP for Sleuth and Rank Products DEGs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Load libraries
```{r}
library(tidyverse)
library(here)
library(clusterProfiler)
library(ggplot2)
library(patchwork)

# You will need the appropriate OrgDb for your organism.
# Magnaporthe oryzae: use org.Moryzae.eg.db if available, otherwise use a custom
# gene2GO mapping. If you have a custom mapping, set use_custom_go = TRUE below.

# ── USER SETTINGS ──────────────────────────────────────────────────────────────
use_custom_go  <- TRUE   # Set FALSE if using an OrgDb package instead
go_mapping_file <- here("Raw", "go_annotations.txt")  
# Expected format: two columns, tab-separated: gene_id | GO_id
# e.g.:  MGG_00001   GO:0003674
#        MGG_00001   GO:0008150

# If use_custom_go = FALSE, set your OrgDb:
# library(org.Moryzae.eg.db)
# orgdb <- org.Moryzae.eg.db

p_cutoff   <- 0.05
q_cutoff   <- 0.05
min_gene_size <- 5   # minimum genes in a GO term to test
# ───────────────────────────────────────────────────────────────────────────────
```

# Output directory
```{r}
fig_dir <- "C:/Users/ges23jir/Documents/Coding/TSL/mst7_transcriptome/Analysis/Data/Figures/DEG_analysis"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

save_plot <- function(plot, filename, width = 10, height = 7) {
  ggsave(file.path(fig_dir, paste0(filename, ".pdf")),
         plot, width = width, height = height, device = "pdf")
  ggsave(file.path(fig_dir, paste0(filename, ".svg")),
         plot, width = width, height = height, device = "svg")
  message("Saved: ", filename)
}
```

# Load GO mapping
```{r}
if (use_custom_go) {
  go_map <- read_tsv(go_mapping_file, col_names = c("gene_id", "go_id"),
                     col_types = cols()) %>%
    filter(!is.na(go_id), !is.na(gene_id))
  
  # term2gene: GO ID -> gene
  term2gene <- go_map %>% select(go_id, gene_id)
  
  # term2name: GO ID -> GO term name (requires GO.db)
  library(GO.db)
  go_terms <- AnnotationDbi::select(GO.db,
                                    keys    = unique(go_map$go_id),
                                    columns = c("GOID", "TERM", "ONTOLOGY"),
                                    keytype = "GOID") %>%
    as_tibble() %>%
    rename(go_id = GOID, go_name = TERM, ontology = ONTOLOGY)
  
  term2name_BP <- go_terms %>% filter(ontology == "BP") %>% select(go_id, go_name)
  term2name_MF <- go_terms %>% filter(ontology == "MF") %>% select(go_id, go_name)
  term2gene_BP  <- term2gene %>% filter(go_id %in% term2name_BP$go_id)
  term2gene_MF  <- term2gene %>% filter(go_id %in% term2name_MF$go_id)
  
  # Universe = all genes with any GO annotation
  go_universe <- unique(go_map$gene_id)
  
  message("GO mapping loaded: ", nrow(go_map), " gene-GO pairs")
  message("Universe size: ", length(go_universe), " genes")
}
```

# Helper: run enrichment for one gene list
```{r}
run_go <- function(gene_list, universe, label,
                   term2gene_bp, term2name_bp,
                   term2gene_mf, term2name_mf) {
  
  if (length(gene_list) < 3) {
    message("  Skipping ", label, ": fewer than 3 genes")
    return(NULL)
  }
  
  run_ora <- function(t2g, t2n, ont_label) {
    tryCatch({
      enricher(
        gene          = gene_list,
        universe      = universe,
        TERM2GENE     = t2g,
        TERM2NAME     = t2n,
        pvalueCutoff  = p_cutoff,
        qvalueCutoff  = q_cutoff,
        minGSSize     = min_gene_size
      )
    }, error = function(e) {
      message("  ORA error (", ont_label, "): ", e$message)
      NULL
    })
  }
  
  list(
    BP = run_ora(term2gene_bp, term2name_bp, "BP"),
    MF = run_ora(term2gene_mf, term2name_mf, "MF")
  )
}
```

# Helper: dotplot for one enrichment result
```{r}
go_dotplot <- function(enrich_obj, title, top_n = 20) {
  if (is.null(enrich_obj) || nrow(as.data.frame(enrich_obj)) == 0) return(NULL)
  
  df <- as.data.frame(enrich_obj) %>%
    arrange(p.adjust) %>%
    slice_head(n = top_n) %>%
    mutate(
      Description  = str_wrap(Description, width = 50),
      GeneRatio_num = sapply(GeneRatio, function(x) {
        parts <- str_split(x, "/")[[1]]
        as.numeric(parts[1]) / as.numeric(parts[2])
      })
    )
  
  ggplot(df, aes(x = GeneRatio_num, y = reorder(Description, GeneRatio_num),
                 colour = p.adjust, size = Count)) +
    geom_point() +
    scale_colour_gradient(low = "#D62728", high = "#AEC7E8",
                          name = "adj. p-value") +
    scale_size_continuous(name = "Gene count", range = c(2, 8)) +
    labs(title = title, x = "Gene Ratio", y = NULL) +
    theme_classic(base_size = 11) +
    theme(
      plot.title  = element_text(face = "bold", size = 10),
      axis.text.y = element_text(size = 8)
    )
}
```

# ── SLEUTH GO ANALYSIS ────────────────────────────────────────────────────────

## Load sleuth results
```{r}
dge_dir     <- here("Analysis/Data/DGE")
sleuth_files <- list.files(dge_dir, pattern = "^sleuth_.*\\.csv$", full.names = TRUE)

sleuth_all <- lapply(sleuth_files, function(f) {
  df <- read_csv(f, col_types = cols())
  bn <- basename(f) %>% gsub("^sleuth_|\\.csv$", "", .)
  parts <- str_split(bn, "_")[[1]]
  df %>% mutate(
    experiment  = parts[1],
    timepoint   = parts[2],
    condition   = parts[3],
    control     = parts[5],
    comparison  = paste0(parts[3], "_vs_", parts[5])
  )
}) %>% bind_rows() %>%
  mutate(
    sig       = !is.na(qval) & qval < 0.05,
    direction = case_when(
      sig & b > 0 ~ "up",
      sig & b < 0 ~ "down",
      TRUE        ~ "ns"
    ),
    timepoint_num = as.numeric(gsub("H", "", timepoint))
  )

# Strip transcript version suffixes if needed (e.g. MGG_00001.1 -> MGG_00001)
# Uncomment and adjust if your IDs have version suffixes:
# sleuth_all <- sleuth_all %>%
#   mutate(gene_id = gsub("\\.\\d+$", "", target_id))
# For now assume target_id matches GO mapping directly:
sleuth_all <- sleuth_all %>% mutate(gene_id = target_id)
```

## Run GO enrichment for each sleuth comparison × timepoint × direction
```{r}
sleuth_comparisons <- unique(sleuth_all$comparison)
sleuth_timepoints  <- unique(sleuth_all$timepoint)

for (comp in sleuth_comparisons) {
  for (tp in sleuth_timepoints) {
    
    df_sub <- sleuth_all %>% filter(comparison == comp, timepoint == tp)
    if (nrow(df_sub) == 0) next
    
    for (dir in c("up", "down")) {
      gene_list <- df_sub %>%
        filter(direction == dir) %>%
        pull(gene_id) %>%
        unique() %>%
        intersect(go_universe)
      
      label <- paste0("sleuth_", comp, "_", tp, "_", dir)
      message("Running GO: ", label, " (n=", length(gene_list), ")")
      
      go_res <- run_go(gene_list, go_universe, label,
                       term2gene_BP, term2name_BP,
                       term2gene_MF, term2name_MF)
      
      # Plot BP
      p_bp <- go_dotplot(go_res$BP,
                         title = paste0(gsub("_vs_", " vs ", comp),
                                        " | ", tp, " | ", dir, "regulated\nBP"))
      # Plot MF
      p_mf <- go_dotplot(go_res$MF,
                         title = paste0(gsub("_vs_", " vs ", comp),
                                        " | ", tp, " | ", dir, "regulated\nMF"))
      
      # Combine BP + MF side by side, skip if both NULL
      plots_available <- list(p_bp, p_mf) %>% Filter(Negate(is.null), .)
      if (length(plots_available) == 0) {
        message("  No significant GO terms for ", label)
        next
      }
      
      combined <- wrap_plots(plots_available, ncol = length(plots_available)) +
        plot_annotation(
          title    = paste0("GO Enrichment: ", gsub("_vs_", " vs ", comp)),
          subtitle = paste0(tp, " | ", dir, "regulated | sleuth | qval < 0.05"),
          theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                           plot.subtitle = element_text(size = 10))
        )
      
      save_plot(combined,
                paste0("GO_sleuth_", comp, "_", tp, "_", dir),
                width  = 7 * length(plots_available),
                height = 8)
      
      # Save enrichment tables
      go_out_dir <- here("Analysis/Data/GO")
      dir.create(go_out_dir, recursive = TRUE, showWarnings = FALSE)
      
      if (!is.null(go_res$BP) && nrow(as.data.frame(go_res$BP)) > 0)
        write_csv(as.data.frame(go_res$BP),
                  file.path(go_out_dir, paste0("GO_BP_sleuth_", comp, "_", tp, "_", dir, ".csv")))
      
      if (!is.null(go_res$MF) && nrow(as.data.frame(go_res$MF)) > 0)
        write_csv(as.data.frame(go_res$MF),
                  file.path(go_out_dir, paste0("GO_MF_sleuth_", comp, "_", tp, "_", dir, ".csv")))
    }
  }
}
```

# ── RANK PRODUCTS GO ANALYSIS ─────────────────────────────────────────────────

## Load rank products results
```{r}
rp_dir   <- here("Analysis/Data/DGE/RankProducts")
rp_files <- list.files(rp_dir, pattern = "^rankprod_.*\\.csv$", full.names = TRUE)

rp_all <- lapply(rp_files, function(f) {
  df  <- read_csv(f, col_types = cols())
  bn  <- basename(f) %>% gsub("^rankprod_|\\.csv$", "", .)
  
  tp_str   <- sub("_.*$", "", bn)
  cond_str <- sub("^[^_]+_", "", bn) %>% sub("_vs_.*$", "", .)
  ctrl_str <- sub("^.*_vs_", "", bn)
  
  # Use CSV columns if present (reverse comparisons write them explicitly)
  if ("condition" %in% names(df) && "control" %in% names(df)) {
    df %>% mutate(
      timepoint  = tp_str,
      comparison = paste0(unique(condition), "_vs_", unique(control))
    )
  } else {
    df %>% mutate(
      timepoint  = tp_str,
      condition  = cond_str,
      control    = ctrl_str,
      comparison = paste0(cond_str, "_vs_", ctrl_str)
    )
  }
}) %>% bind_rows() %>%
  mutate(
    sig           = !is.na(pfp) & pfp < 0.05,
    timepoint_num = as.numeric(gsub("H", "", timepoint))
  )

# Harmonise gene IDs if needed (same as sleuth)
rp_all <- rp_all %>% mutate(gene_id = target_id)
```

## Run GO enrichment for each rank products comparison × timepoint × direction
```{r}
rp_comparisons <- unique(rp_all$comparison)
rp_timepoints  <- unique(rp_all$timepoint)

for (comp in rp_comparisons) {
  grp1 <- unique(rp_all$condition[rp_all$comparison == comp])
  
  for (tp in rp_timepoints) {
    
    df_sub <- rp_all %>% filter(comparison == comp, timepoint == tp, sig)
    if (nrow(df_sub) == 0) next
    
    # Use exact match on "up_in_<grp1>" to avoid Guy11 matching inside Guy11M
    df_sub <- df_sub %>%
      mutate(dir_simple = if_else(direction == paste0("up_in_", grp1), "up", "down"))
    
    for (dir in c("up", "down")) {
      gene_list <- df_sub %>%
        filter(dir_simple == dir) %>%
        pull(gene_id) %>%
        unique() %>%
        intersect(go_universe)
      
      label <- paste0("rankprod_", comp, "_", tp, "_", dir)
      message("Running GO: ", label, " (n=", length(gene_list), ")")
      
      go_res <- run_go(gene_list, go_universe, label,
                       term2gene_BP, term2name_BP,
                       term2gene_MF, term2name_MF)
      
      p_bp <- go_dotplot(go_res$BP,
                         title = paste0(gsub("_vs_", " vs ", comp),
                                        " | ", tp, " | ", dir, "regulated\nBP"))
      p_mf <- go_dotplot(go_res$MF,
                         title = paste0(gsub("_vs_", " vs ", comp),
                                        " | ", tp, " | ", dir, "regulated\nMF"))
      
      plots_available <- list(p_bp, p_mf) %>% Filter(Negate(is.null), .)
      if (length(plots_available) == 0) {
        message("  No significant GO terms for ", label)
        next
      }
      
      combined <- wrap_plots(plots_available, ncol = length(plots_available)) +
        plot_annotation(
          title    = paste0("GO Enrichment: ", gsub("_vs_", " vs ", comp)),
          subtitle = paste0(tp, " | ", dir, "regulated | Rank Products | pfp < 0.05"),
          theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                           plot.subtitle = element_text(size = 10))
        )
      
      save_plot(combined,
                paste0("GO_rankprod_", comp, "_", tp, "_", dir),
                width  = 7 * length(plots_available),
                height = 8)
      
      # Save tables
      go_out_dir <- here("Analysis/Data/GO")
      dir.create(go_out_dir, recursive = TRUE, showWarnings = FALSE)
      
      if (!is.null(go_res$BP) && nrow(as.data.frame(go_res$BP)) > 0)
        write_csv(as.data.frame(go_res$BP),
                  file.path(go_out_dir, paste0("GO_BP_rankprod_", comp, "_", tp, "_", dir, ".csv")))
      
      if (!is.null(go_res$MF) && nrow(as.data.frame(go_res$MF)) > 0)
        write_csv(as.data.frame(go_res$MF),
                  file.path(go_out_dir, paste0("GO_MF_rankprod_", comp, "_", tp, "_", dir, ".csv")))
    }
  }
}

message("All GO plots and tables saved.")
```
