---
title: "Modular RNA-seq pipeline for mst7 transcriptome"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Overview

This document runs the modular RNA-seq workflow for the mst7 transcriptome project. It reads the sample metadata, sources the reusable analysis scripts, runs sleuth-based DGE comparisons from the comparison table, generates DEG summary plots, creates upset plots, and generates correlation matrices.

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
library(here)
library(ggplot2)
library(reshape2)
```

```{r paths}
project_dir <- here::here()
analysis_dir <- file.path(project_dir, "analysis")
scripts_dir <- file.path(analysis_dir, "scripts")
results_dir <- file.path(project_dir, "results")
fig_dir <- file.path(results_dir, "figures")
tables_dir <- file.path(results_dir, "tables")
dge_dir <- file.path(results_dir, "dge")
upset_dir <- file.path(fig_dir, "upset")
cor_dir <- file.path(fig_dir, "correlation")

for (dir_path in c(fig_dir, tables_dir, dge_dir, upset_dir, cor_dir)) {
  dir.create(dir_path, recursive = TRUE, showWarnings = FALSE)
}
```

```{r source-scripts}
source(file.path(scripts_dir, "comparison_table.R"))
source(file.path(scripts_dir, "upset_functions.R"))
source(file.path(scripts_dir, "correlation_functions.R"))
source(file.path(scripts_dir, "dge_functions.R"))
source(file.path(scripts_dir, "rank_products_functions.R"))
```

```{r metadata}
s2c <- read_csv(file.path(project_dir, "raw", "run_metadata.txt"), show_col_types = FALSE) %>%
  mutate(
    path = file.path(project_dir, "raw", path),
    name = str_remove(name, "_0h_\\d+$"),
    sample = paste0(name, "_", timepoint, "_", biorep),
    name = factor(name),
    timepoint = factor(timepoint),
    biorep = factor(biorep)
  ) %>%
  mutate(
    experiment = if_else(name %in% c("pmk1", "Guy11M"), "Exp2", "Exp1"),
    experiment = factor(experiment, levels = c("Exp1", "Exp2"))
  )
```

# Sleuth differential expression

```{r run-dge, message=FALSE, warning=FALSE}
dge_results <- run_comparisons(s2c, comparisons, out_dir = dge_dir)
```

```{r dge-summary-plots}
for (lab in unique(dge_results$comparison_label)) {
  if (!is.na(lab) && nzchar(lab)) {
    plot_dge_summary(dge_results, lab, out_dir = fig_dir)
  }
}
```

# Upset plots

```{r upset-plots}
deg_sets <- load_deg_tables(
  comparison_labels = comparisons$label,
  dge_dir = dge_dir,
  qval_threshold = 0.05,
  fc_threshold = 0
)

plot_upset(deg_sets, label = "all_comparisons", out_dir = upset_dir)
```

# Correlation matrices

```{r correlation-matrices}
tpm_matrix <- load_tpm_matrix(file.path(project_dir, "raw", "all_samples_tpm_matrix.txt"))

cor_matrix_all(tpm_matrix, s2c)
cor_matrix_exp1(tpm_matrix, s2c)
cor_matrix_exp2(tpm_matrix, s2c)
cor_matrix_key_strains(tpm_matrix, s2c)
```

# Rank-products plots

```{r rank-products}
rp_results <- run_rank_products_comparisons(s2c, comparisons, out_dir = file.path(results_dir, "rank_products"))

for (lab in unique(rp_results$comparison_label)) {
  if (!is.na(lab) && nzchar(lab)) {
    plot_rank_products_mirrored(
      rp_results,
      lab,
      out_dir = file.path(fig_dir, "rank_products")
    )
  }
}
```

# Session info

```{r session-info}
sessionInfo()
```
