---
title: "Cross-Sample Correlation Matrix from Kallisto TPM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, 
                      fig.width = 12, fig.height = 10)
```

# Load Libraries
```{r}
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(corrplot)
library(here)

# Create output directory
dir.create(here("Analysis/Data/tpm_correlation_analysis"), recursive = TRUE, showWarnings = FALSE)
```

# 1. Load Sample Metadata

```{r}
# Load the sample metadata
sample_metadata <- read_csv(here("Raw/results/run_metadata.txt"), show_col_types = FALSE)

# Create a clean sample ID for labeling
sample_metadata <- sample_metadata %>%
  mutate(
    sample_id = paste(name, timepoint, biorep, sep = "_"),
    timepoint_label = paste0(timepoint, "H"),
    path = file.path("Raw/results", path)
  )

cat("Total samples:", nrow(sample_metadata), "\n")
print(sample_metadata)
```

# 2. Load TPM Data from Kallisto

```{r}
# Function to load kallisto abundance.tsv file
load_kallisto_tpm <- function(file_path, sample_name) {
  if (file.exists(file_path)) {
    df <- read_tsv(file_path, show_col_types = FALSE) %>%
      select(target_id, tpm) %>%
      rename(!!sample_name := tpm)
    return(df)
  } else {
    warning(paste("File not found:", file_path))
    return(NULL)
  }
}

# Load all TPM data
tpm_list <- list()

for (i in 1:nrow(sample_metadata)) {
  sample_id <- sample_metadata$sample_id[i]
  file_path <- here(sample_metadata$path[i])
  
  tpm_data <- load_kallisto_tpm(file_path, sample_id)
  
  if (!is.null(tpm_data)) {
    tpm_list[[sample_id]] <- tpm_data
    cat("Loaded:", sample_id, "from", file_path, "\n")
  }
}

# Check how many samples were loaded
cat("\nSuccessfully loaded", length(tpm_list), "out of", nrow(sample_metadata), "samples\n")

# If no samples loaded, print error
if (length(tpm_list) == 0) {
  stop("!!! NO SAMPLES LOADED !!! Please check your file paths.")
}
```

# 3. Create TPM Matrix

```{r}
# Merge all TPM data into a single wide matrix
tpm_matrix <- tpm_list %>%
  reduce(full_join, by = "target_id") %>%
  column_to_rownames("target_id") %>%
  as.matrix()

cat("TPM matrix dimensions:", nrow(tpm_matrix), "genes x", ncol(tpm_matrix), "samples\n")
cat("First few rows and columns:\n")
print(tpm_matrix[1:5, 1:min(5, ncol(tpm_matrix))])

# Save the TPM matrix
write.csv(tpm_matrix, here("Analysis/Data/tpm_matrix_all_samples.csv"))
```

# 4. Filter Low Expression Genes (Optional but Recommended)

```{r}
# Filter genes with very low expression across all samples
# This reduces noise in correlation analysis

# Keep genes with mean TPM > 1
mean_tpm <- rowMeans(tpm_matrix, na.rm = TRUE)
tpm_filtered <- tpm_matrix[mean_tpm > 1, ]

cat("After filtering (mean TPM > 1):\n")
cat("Genes retained:", nrow(tpm_filtered), "out of", nrow(tpm_matrix), "\n")
cat("Percentage retained:", round(100 * nrow(tpm_filtered) / nrow(tpm_matrix), 1), "%\n")
```

# 5. Log-Transform TPM Values

```{r}
# Log2 transformation (add pseudocount to avoid log(0))
tpm_log2 <- log2(tpm_filtered + 1)

cat("Log2-transformed TPM matrix dimensions:", 
    nrow(tpm_log2), "genes x", ncol(tpm_log2), "samples\n")
```
# 6. Calculate Correlation Matrix

```{r}
# Calculate Pearson correlation between all samples
cor_matrix_pearson <- cor(tpm_log2, use = "pairwise.complete.obs", method = "pearson")

# Calculate Spearman correlation
cor_matrix_spearman <- cor(tpm_log2, use = "pairwise.complete.obs", method = "spearman")

cat("\n=== PEARSON CORRELATION MATRIX (first 5x5) ===\n")
print(round(cor_matrix_pearson[1:5, 1:5], 3))

cat("\n=== SPEARMAN CORRELATION MATRIX (first 5x5) ===\n")
print(round(cor_matrix_spearman[1:5, 1:5], 3))

# Calculate summary statistics
cat("\n=== CORRELATION SUMMARY STATISTICS ===\n")
cat("Pearson correlations:\n")
cat("  Mean:", round(mean(cor_matrix_pearson[upper.tri(cor_matrix_pearson)]), 3), "\n")
cat("  Median:", round(median(cor_matrix_pearson[upper.tri(cor_matrix_pearson)]), 3), "\n")
cat("  Range:", round(min(cor_matrix_pearson[upper.tri(cor_matrix_pearson)]), 3), "-", 
    round(max(cor_matrix_pearson[upper.tri(cor_matrix_pearson)]), 3), "\n")

# Save correlation matrices
write.csv(cor_matrix_pearson, 
          here("Analysis/Data/correlation_matrix_pearson_tpm.csv"))
write.csv(cor_matrix_spearman, 
          here("Analysis/Data/correlation_matrix_spearman_tpm.csv"))
```

# 7. Create Annotation for Heatmap

```{r}
# Create annotation dataframe from sample information
annotation_df <- sample_metadata %>%
  filter(sample_id %in% colnames(cor_matrix_pearson)) %>%
  select(sample_id, name, timepoint_label) %>%
  column_to_rownames("sample_id")

# Reorder to match correlation matrix
annotation_df <- annotation_df[colnames(cor_matrix_pearson), ]

# Define colors for annotation (no strain_group)
ann_colors <- list(
  name = c(
    Guy11 = "#1b9e77", 
    Guy11M = "#d95f02",
    mst7 = "#7570b3", 
    pmk1 = "#e7298a",
    MST7WT = "#66a61e",
    MST7PM = "#e6ab02",
    MST7PD = "#a6761d"
  ),
  timepoint_label = c(
    "0H" = "#ffffcc", 
    "2H" = "#a1dab4", 
    "4H" = "#41b6c4", 
    "24H" = "#225ea8"
  )
)

cat("Annotation prepared for", nrow(annotation_df), "samples\n")
```

# 8. Visualize Correlation Matrix - Heatmap

```{r, fig.width=16, fig.height=14}
# Pearson correlation heatmap
pheatmap(
  cor_matrix_pearson,
  color = colorRampPalette(c("#313695", "#4575b4", "#74add1", 
                             "#abd9e9", "#e0f3f8", "#ffffbf",
                             "#fee090", "#fdae61", "#f46d43", 
                             "#d73027", "#a50026"))(100),
  breaks = seq(0.5, 1, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_row = annotation_df,
  annotation_col = annotation_df,
  annotation_colors = ann_colors,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize = 7,
  fontsize_row = 6,
  fontsize_col = 6,
  main = "Sample Correlation Matrix - Pearson (Log2 TPM)",
  filename = here("Analysis/Data/Figures/correlation_heatmap_pearson_tpm.pdf"),
  width = 16,
  height = 14
)

# Spearman correlation heatmap
pheatmap(
  cor_matrix_spearman,
  color = colorRampPalette(c("#313695", "#4575b4", "#74add1", 
                             "#abd9e9", "#e0f3f8", "#ffffbf",
                             "#fee090", "#fdae61", "#f46d43", 
                             "#d73027", "#a50026"))(100),
  breaks = seq(0.5, 1, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_row = annotation_df,
  annotation_col = annotation_df,
  annotation_colors = ann_colors,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize = 7,
  fontsize_row = 6,
  fontsize_col = 6,
  main = "Sample Correlation Matrix - Spearman (Log2 TPM)",
  filename = here("Analysis/Data/Figures/correlation_heatmap_spearman_tpm.pdf"),
  width = 16,
  height = 14
)

cat("\nCorrelation heatmaps saved!\n")
```

# 9. Alternative Visualization - Corrplot

```{r, fig.width=16, fig.height=16}
# Create a correlation plot with corrplot package
pdf(here("Analysis/Data/Figures/corrplot_pearson_tpm.pdf"), width = 16, height = 16)
corrplot(
  cor_matrix_pearson, 
  method = "color",
  type = "full",
  order = "hclust",
  tl.col = "black",
  tl.cex = 0.6,
  cl.cex = 0.8,
  col = colorRampPalette(c("#313695", "#4575b4", "#abd9e9", 
                           "#ffffbf", "#fdae61", "#d73027", 
                           "#a50026"))(200),
  title = "Sample Correlation - Pearson (Log2 TPM)",
  mar = c(0,0,2,0)
)
dev.off()

pdf(here("Analysis/Data/Figures/corrplot_spearman_tpm.pdf"), width = 16, height = 16)
corrplot(
  cor_matrix_spearman, 
  method = "color",
  type = "full",
  order = "hclust",
  tl.col = "black",
  tl.cex = 0.6,
  cl.cex = 0.8,
  col = colorRampPalette(c("#313695", "#4575b4", "#abd9e9", 
                           "#ffffbf", "#fdae61", "#d73027", 
                           "#a50026"))(200),
  title = "Sample Correlation - Spearman (Log2 TPM)",
  mar = c(0,0,2,0)
)
dev.off()

cat("Corrplot visualizations saved!\n")
```

# 10. Correlation Statistics by Group

```{r}
# Create pairwise comparison dataframe
correlation_stats <- expand.grid(
  sample1 = colnames(cor_matrix_pearson),
  sample2 = colnames(cor_matrix_pearson),
  stringsAsFactors = FALSE
) %>%
  filter(sample1 != sample2) %>%
  left_join(sample_metadata %>% select(sample_id, name, timepoint, biorep), 
            by = c("sample1" = "sample_id")) %>%
  rename(name1 = name, time1 = timepoint, rep1 = biorep) %>%
  left_join(sample_metadata %>% select(sample_id, name, timepoint, biorep), 
            by = c("sample2" = "sample_id")) %>%
  rename(name2 = name, time2 = timepoint, rep2 = biorep) %>%
  mutate(
    pearson_cor = map2_dbl(sample1, sample2, ~cor_matrix_pearson[.x, .y]),
    spearman_cor = map2_dbl(sample1, sample2, ~cor_matrix_spearman[.x, .y]),
    comparison_type = case_when(
      name1 == name2 & time1 == time2 & rep1 != rep2 ~ "Biological replicates",
      name1 == name2 & time1 != time2 ~ "Same strain, different time",
      name1 != name2 & time1 == time2 ~ "Different strain, same time",
      TRUE ~ "Different strains, different time"
    )
  )

# Summary statistics
correlation_summary <- correlation_stats %>%
  group_by(comparison_type) %>%
  summarise(
    n_comparisons = n(),
    mean_pearson = mean(pearson_cor, na.rm = TRUE),
    median_pearson = median(pearson_cor, na.rm = TRUE),
    sd_pearson = sd(pearson_cor, na.rm = TRUE),
    min_pearson = min(pearson_cor, na.rm = TRUE),
    max_pearson = max(pearson_cor, na.rm = TRUE),
    mean_spearman = mean(spearman_cor, na.rm = TRUE),
    median_spearman = median(spearman_cor, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_pearson))

cat("\n=== CORRELATION SUMMARY BY COMPARISON TYPE ===\n")
print(correlation_summary)

write_csv(
  correlation_summary, 
  here("Analysis/Data/correlation_summary_by_type_tpm.csv")
)
write_csv(
  correlation_stats,
  here("Analysis/Data/pairwise_correlations_detailed_tpm.csv")
)
```

# 11. Visualize Correlation Distribution

```{r, fig.width=14, fig.height=8}
library(ggrepel)

# Create labels for points
correlation_stats <- correlation_stats %>%
  mutate(point_label = paste0(sample1, " vs ", sample2))

# Boxplot of correlations by comparison type with labeled points
ggplot(correlation_stats, aes(x = reorder(comparison_type, -pearson_cor), 
                              y = pearson_cor, fill = comparison_type)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  geom_text_repel(
    aes(label = point_label),
    size = 2,
    max.overlaps = 20,
    box.padding = 0.3,
    point.padding = 0.2,
    segment.size = 0.2,
    segment.alpha = 0.5
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Sample Correlations (Pearson) - Log2 TPM Data",
    subtitle = paste0("Based on ", nrow(tpm_log2), " filtered genes (mean TPM > 1)"),
    x = "Comparison Type",
    y = "Pearson Correlation",
    fill = "Comparison Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  coord_cartesian(ylim = c(0, 1))

ggsave(here("Analysis/Data/Figures/correlation_distribution_boxplot_tpm.pdf"),
       width = 14, height = 8)

# Save a version with fewer labels (only outliers and extreme values)
correlation_stats_labeled <- correlation_stats %>%
  group_by(comparison_type) %>%
  mutate(
    is_outlier = pearson_cor < (quantile(pearson_cor, 0.25) - 1.5 * IQR(pearson_cor)) |
                 pearson_cor > (quantile(pearson_cor, 0.75) + 1.5 * IQR(pearson_cor)),
    is_extreme = pearson_cor < quantile(pearson_cor, 0.05) | 
                 pearson_cor > quantile(pearson_cor, 0.95)
  ) %>%
  ungroup()

ggplot(correlation_stats_labeled, aes(x = reorder(comparison_type, -pearson_cor), 
                                       y = pearson_cor, fill = comparison_type)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  geom_text_repel(
    data = correlation_stats_labeled %>% filter(is_outlier | is_extreme),
    aes(label = point_label),
    size = 2.5,
    max.overlaps = 30,
    box.padding = 0.5,
    point.padding = 0.3,
    segment.size = 0.3,
    segment.alpha = 0.6,
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Sample Correlations (Pearson) - Outliers Labeled",
    subtitle = paste0("Based on ", nrow(tpm_log2), " filtered genes (mean TPM > 1)"),
    x = "Comparison Type",
    y = "Pearson Correlation",
    fill = "Comparison Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  coord_cartesian(ylim = c(0, 1))

ggsave(here("Analysis/Data/Figures/correlation_distribution_boxplot_outliers_labeled.pdf"),
       width = 14, height = 8)

# Boxplot of correlations by comparison type WITHOUT labels
ggplot(correlation_stats, aes(x = reorder(comparison_type, -pearson_cor), 
                              y = pearson_cor, fill = comparison_type)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Sample Correlations (Pearson) - No Labels",
    subtitle = paste0("Based on ", nrow(tpm_log2), " filtered genes (mean TPM > 1)"),
    x = "Comparison Type",
    y = "Pearson Correlation",
    fill = "Comparison Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  coord_cartesian(ylim = c(0, 1))

ggsave(here("Analysis/Data/Figures/correlation_distribution_boxplot_tpm_nolabels.pdf"),
       width = 14, height = 8)
```

# 12. Replicate Quality Check

```{r, fig.width=14, fig.height=8}
library(ggrepel)

# Extract correlations between biological replicates
replicate_cors <- correlation_stats %>%
  filter(comparison_type == "Biological replicates") %>%
  # Remove duplicate comparisons by keeping only one direction
  # Sort sample names alphabetically and keep first occurrence
  rowwise() %>%
  mutate(
    sample_min = min(sample1, sample2),
    sample_max = max(sample1, sample2),
    comparison_key = paste(sample_min, sample_max, sep = "_vs_")
  ) %>%
  ungroup() %>%
  distinct(comparison_key, .keep_all = TRUE) %>%
  select(-sample_min, -sample_max, -comparison_key) %>%
  arrange(pearson_cor) %>%
  mutate(
    pair_label = paste(sample1, "vs", sample2),
    # Create a shorter label showing strain and timepoint
    short_label = paste0(name1, "_", time1, "H (r", rep1, " vs r", rep2, ")")
  )

cat("\n=== BIOLOGICAL REPLICATE CORRELATIONS ===\n")
cat("Number of unique replicate pairs:", nrow(replicate_cors), "\n")
cat("Mean correlation:", round(mean(replicate_cors$pearson_cor), 3), "\n")
cat("Median correlation:", round(median(replicate_cors$pearson_cor), 3), "\n")
cat("Min correlation:", round(min(replicate_cors$pearson_cor), 3), "\n")
cat("Max correlation:", round(max(replicate_cors$pearson_cor), 3), "\n\n")

# Check for any low-quality replicates (correlation < 0.9)
low_cor_reps <- replicate_cors %>%
  filter(pearson_cor < 0.9)

if (nrow(low_cor_reps) > 0) {
  cat("WARNING: Found", nrow(low_cor_reps), "replicate pairs with correlation < 0.9:\n")
  print(low_cor_reps %>% select(sample1, sample2, pearson_cor))
} else {
  cat("All biological replicates have correlation >= 0.9\n")
}

# Visualize replicate correlations with all points labeled
ggplot(replicate_cors, aes(x = reorder(short_label, pearson_cor), y = pearson_cor)) +
  geom_point(aes(color = pearson_cor), size = 3) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "blue", linewidth = 1) +
  geom_text_repel(
    aes(label = short_label),
    size = 2.5,
    max.overlaps = Inf,
    box.padding = 0.5,
    point.padding = 0.3,
    segment.size = 0.2,
    segment.alpha = 0.5,
    direction = "both",
    hjust = 0
  ) +
  scale_color_gradient2(
    low = "#d73027",
    mid = "#fee090",
    high = "#1a9850",
    midpoint = 0.95,
    name = "Correlation"
  ) +
  labs(
    title = "Biological Replicate Correlations",
    subtitle = "Dashed line indicates correlation = 0.9 (duplicate pairs removed)",
    x = "",
    y = "Pearson Correlation"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  ) +
  coord_cartesian(ylim = c(0.5, 1))

ggsave(here("Analysis/Data/Figures/replicate_correlations_labeled.pdf"), 
       width = 14, height = 8)

# Create an alternative horizontal version for better readability
ggplot(replicate_cors, aes(y = reorder(short_label, pearson_cor), x = pearson_cor)) +
  geom_point(aes(color = pearson_cor), size = 3) +
  geom_vline(xintercept = 0.9, linetype = "dashed", color = "blue", linewidth = 1) +
  geom_segment(aes(x = 0.5, xend = pearson_cor, 
                   y = short_label, yend = short_label, 
                   color = pearson_cor), 
               linewidth = 1, alpha = 0.5) +
  scale_color_gradient2(
    low = "#d73027",
    mid = "#fee090",
    high = "#1a9850",
    midpoint = 0.95,
    name = "Correlation"
  ) +
  labs(
    title = "Biological Replicate Correlations",
    subtitle = "Dashed line indicates correlation = 0.9 (duplicate pairs removed)",
    y = "Sample Pair",
    x = "Pearson Correlation"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  ) +
  coord_cartesian(xlim = c(0.5, 1))

ggsave(here("Analysis/Data/Figures/replicate_correlations_horizontal.pdf"), 
       width = 12, height = max(8, nrow(replicate_cors) * 0.2))

# Create a detailed table of replicate correlations
replicate_table <- replicate_cors %>%
  select(sample1, sample2, name1, time1, rep1, rep2, pearson_cor, spearman_cor) %>%
  arrange(pearson_cor)

write_csv(
  replicate_table,
  here("Analysis/Data/replicate_correlations_table.csv")
)

cat("\nReplicate correlation table saved!\n")
```

# 13. PCA Plot for Visual Quality Check

```{r, fig.width=12, fig.height=8}
# Perform PCA on log2-transformed TPM
pca_result <- prcomp(t(tpm_log2), scale. = TRUE)

# Extract PC scores
pca_data <- as.data.frame(pca_result$x[, 1:5]) %>%
  rownames_to_column("sample_id") %>%
  left_join(sample_metadata, by = "sample_id")

# Calculate variance explained
var_explained <- summary(pca_result)$importance[2, ] * 100

# PCA plot - PC1 vs PC2
p1 <- ggplot(pca_data, aes(x = PC1, y = PC2, color = name, shape = timepoint_label)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = ann_colors$name) +
  labs(
    title = "PCA of Log2 TPM Expression Data",
    x = paste0("PC1 (", round(var_explained[1], 1), "% variance)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "% variance)"),
    color = "Strain",
    shape = "Timepoint"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

print(p1)
ggsave(here("Analysis/Data/Figures/PCA_plot_PC1_PC2.pdf"),
       width = 12, height = 8)

# PCA plot - PC2 vs PC3
p2 <- ggplot(pca_data, aes(x = PC2, y = PC3, color = name, shape = timepoint_label)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = ann_colors$name) +
  labs(
    title = "PCA of Log2 TPM Expression Data",
    x = paste0("PC2 (", round(var_explained[2], 1), "% variance)"),
    y = paste0("PC3 (", round(var_explained[3], 1), "% variance)"),
    color = "Strain",
    shape = "Timepoint"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

print(p2)
ggsave(here("Analysis/Data/Figures/PCA_plot_PC2_PC3.pdf"),
       width = 12, height = 8)

cat("\nPCA analysis complete!\n")
cat("PC1 explains", round(var_explained[1], 1), "% of variance\n")
cat("PC2 explains", round(var_explained[2], 1), "% of variance\n")
cat("PC3 explains", round(var_explained[3], 1), "% of variance\n")
```

# 14. Session Info

```{r}
sessionInfo()
```
